A geometric approach to linear and quadratic Chabauty

### 2020-02-27 MS.tex
Finished theoretical chapter

\subsection{From \texorpdfstring{$C(\Z_p)$}{C(Z\_p)} to \texorpdfstring{$J(\Z_p)$}{J(Z\_p)}}
\label{subsection:czptojzp}
We know that $C(\Z_p)_P$ is in bijection with $\Z_p$, again with the bijection given by evaluating a parameter and dividing by $p$. The resulting function $\Z_p \to \Z_p^g$ is linear and non-constant modulo $p$, i.e. there are power series $f_1,\dots,f_{g-1} \in \Z_p\angle{z_1,...,z_g}$ such that the image of $C(\Z_p)_P$ is exactly given by $V(f_1,...,f_{g-1})$, and all $f_i$ are linear modulo $p$. Another way to think of this, is as $C(\Z/p^2\Z)_P$ being an affine line inside $J(\Z/p^2\Z)_t$.
TODO: give proof. This uses the functoriality of smooth $\Z_p$-points reducing to a point. That $C(\Z/p^2\Z)_P$ is an affine line inside $J(\Z/p^2\Z)_t$ can also be seen more easily, using Lemma~\ref{lem:smoothpointsfunc}. 
% Say something about the f_i defining C(Z_p) in J(Z_p)

@@

\section{Computing the intersection}
\label{section:intersection}
Clearly, as subsets of $J(\Z_p)_t$, we have the inclusion $C(\Z)_P \subset C(\Z_p)_P \cap \overline{J(\Z)_t}$. The theory we have built so far enables the following method, which is in sharp contrast with Colemans method; instead of pulling back equations for $\overline{J(\Z)_t}$ to $C(\Z_p)_P$, we pull back equations for $C(\Z_p)_P$ to $\overline{J(\Z)_t}$ to arrive at the following theorem.
\begin{thm}
Let $C,J,p,P,t$ be as in the overview, let $\kappa$ be as in Theorem~\ref{thm:kappanice} and let $f_1,\cdots,f_{g-1}$ be as in Subsection~\ref{subsection:czptojzp}. Define $\lambda_1 = \kappa^*f_1,\dots,\lambda_{g-1} = \kappa^*f_{g-1}$ to be the pullbacks along $\kappa$ of the $f_i$. Then $\kappa$ induces a surjection from $Z(\lambda_1,\dots,\lambda_{g-1})$ to $C(\Z_p)_P \cap \overline{J(\Z)_t}$.
\end{thm}

The $\lambda_i$ are in general not computationally available, as they are power series, but we can approximate them modulo powers of $p$. For example, we know that the $\lambda_i$ are linear modulo $p$. And in particular, if $f_1,\dots,f_{g-1}$ are as in Remark~\ref{rem:fislinear}, then $\lambda_i = \kappa_i$ for $i \leq g-1$, so they are also of degree at most $m$ modulo $p^m$ for $m < p-1$. Knowing this, we can compute $\lambda_i$ modulo powers of $p$ by extrapolation, and the following lemma, an immediate consequence of Lagrange interpolation, tells us how.
\begin{lem}
\label{lem:interpolate}
Let $R$ be a ring, and assume operations in $R$ are $O(1)$, $f \in R[x]$ with $x = (x_1,\dots,x_n)$ a polynomial of degree at most $m$, and $S$ a subset of $R$ with $S-S \subset \{0\} \cup R^*$ and $|S| > m$. Then $f$ is uniquely determinable in $O(nm^{n+1})$ by its values on $S^n$.
\end{lem}
\begin{proof}
Both uniqueness and construction follow from the case with $n=1$; uniqueness from the fact that the ideals $\{(x-s) \mid s \in S\}$ are pairwise coprime, and the construction is Lagrange interpolation. The complexity follows easily from induction on $n$ and the formula for Lagrange interpolation.
\end{proof}

[TODO: explain how to use these power series. Explain 4.12, $(I : p^\infty)$ and Newton polygons.]
By Theorem 4.12 of Edixhoven-Lido, we just need to check whether $\overline{A}/\overline{I}$ is finite. As all $\overline{\kappa^* f_i}$ are linear, in fact, $\overline{A}/\overline{I}$ is always of the form $0$ (meaning no solutions) or $\F_p[w_1,...,w_s]$ for some $s \leq r$, and $s = 0$ iff the system of linear equations $\overline{\kappa^* f_i} = 0, i \in \{1,...,g-1\}$ has a unique solution. In general, we may hope an upper bound for $|C(\Z)_P|$ of $1$ if $r \leq g-1$.

@@

We first calculate $\Lcal(2D_0 - D_{\text{aff}})$. Both over $\Q_p$ and over $\F_p$ we have the equality
\[
\Lcal(2D_0 - D_{\text{aff}) = a(x)\Lcal((2g+2-d)(\infty_- + \infty_+)) + (y-b(x))\Lcal((g+1)(\infty_- + \infty_+))
\]
and then it should follow over $\Z_p$ so over $\Z/p^k\Z$. \todo{Give more details, currently talking with Mascot about this.}

@@

\subsection{Parameters at $J$}
\label{subs:paramj}
Being able to calculate in the Jacobian, we can move on to the final ingredient for explicit computations: parameters at points of $J$. For any nice curve $C$ over $\Z$, smooth over $\Z_{(p)}$, with a $\Z$-point $b$, we have a birational map from $C^{(g)}$, the $g$-fold symmetric product of $C$, to $J$, given on points by sending $[(P_1,\dots,P_g)]$ to $[\sum P_i - gb]$. This map is \'etale at $[(P_1,\dots,P_g)]$ if $H^0(P_1 + \cdots + P_g)$ has dimension $1$ \todo{find a reference for this}, and if all $P_i$ are distinct, then the map $C^g \to C^{(g)}$ is \'etale at $P_1,\dots,P_g$ as well. Then finding parameters $t_1,\dots,t_g$ at $Q := [\sum_{i=1}^g P_i - gb]$ just comes down to finding parameters at each of the $P_i$. In our case of a hyperellptic curve, this is just an easy computation; for a point $(a,b)$ on the first affine chart, we can take $x-a$ if $b$ is non-zero and $y$ if $a$ is zero.

This now means $J(\Z_p)_Q$ is parametrised as the product of $C(\Z_p)_{P_i}$. In particular, we have a bijection
\[
\prod_{i=1}^g C(\Z/p^k\Z)_{P_i} \to J(\Z/p^k\Z)_{Q}.
\] As we are able to explicitly compute this map using the Abel-Jacobi map, and we are able to test equality in $J(\Z/p^k\Z)$, we can compute the map $(t_1,\dots,t_g): J(\Z/p^k\Z)_Q \to (\Z/p^{k-1}\Z)^g$.

>>>>

\subsection{From \texorpdfstring{$C(\Z_p)$}{C(Z\_p)} to \texorpdfstring{$J(\Z_p)$}{J(Z\_p)}}
\label{subsection:czptojzp}
We know that $C(\Z_p)_P$ is in bijection with $\Z_p$, again with the bijection given by evaluating a parameter and dividing by $p$. The resulting function $\Z_p \to \Z_p^g$ is linear and non-constant modulo $p$, i.e. there are power series $f_1,\dots,f_{g-1} \in \Z_p\angle{z_1,...,z_g}$ such that the image of $C(\Z_p)_P$ is exactly given by $V(f_1,...,f_{g-1})$, and all $f_i$ are linear modulo $p$. Another way to think of this, is as $C(\Z/p^2\Z)_P$ being an affine line inside $J(\Z/p^2\Z)_t$.
\todo{give proof}. This uses the functoriality of smooth $\Z_p$-points reducing to a point. That $C(\Z/p^2\Z)_P$ is an affine line inside $J(\Z/p^2\Z)_t$ can also be seen more easily, using Lemma~\ref{lem:smoothpointsfunc}. 
% Say something about the f_i defining C(Z_p) in J(Z_p)

@@

\section{Computing the intersection}
\label{section:intersection}
Clearly, as subsets of $J(\Z_p)_t$, we have the inclusion $C(\Z)_P \subset C(\Z_p)_P \cap \overline{J(\Z)_t}$. The theory we have built so far enables the following method, which is in sharp contrast with Colemans method; instead of pulling back equations for $\overline{J(\Z)_t}$ to $C(\Z_p)_P$, we pull back equations for $C(\Z_p)_P$ to $\overline{J(\Z)_t}$ to arrive at the following theorem.
\begin{thm}
\label{thm:final}
Let $C,J,p,P,t$ be as in the overview, let $\kappa$ be as in Theorem~\ref{thm:kappanice} and let $f_1,\cdots,f_{g-1}$ be as in Subsection~\ref{subsection:czptojzp}. Define $\lambda_1 = \kappa^*f_1,\dots,\lambda_{g-1} = \kappa^*f_{g-1}$ to be the pullbacks along $\kappa$ of the $f_i$. Then $\kappa$ induces a surjection from $Z(\lambda_1,\dots,\lambda_{g-1}) \subset \Z_p^r$ to $C(\Z_p)_P \cap \overline{J(\Z)_t}$.
\end{thm}

Hence if we find an upper bound for the cardinality of $Z(\lambda_1,\dots,\lambda_{g-1})$, this is also an upper bound for $C(\Z_p)_P$. There are several, sometimes ad hoc, ways of proving finiteness. We let $A$ denote $\Z_p\angle{x_1,\dots,x_r}$ and $I$ the ideal generated by $\lambda_1,\dots,\lambda_{g-1}$. Noting that $Z(\lambda_1,\dots,\lambda_{g-1}) = \Hom_{\Z_p}(A/I,\Z_p)$, we can use the following theorem.

\begin{prop}
\label{prop:finedix}
Let $\overline{A} = A/pA$ and $\overline{I} = I\overline{A}$. Assume $\overline{A}/\overline{I}$ is finite. Then ...% $A$ is the product $\prod_{m \in \MaxSpec(A)} A_m$ and $$\Hom(A/I,\Z_p)$
\end{prop}
\begin{proof}
\todo{give exact statement and proof}
\end{proof}

To utilise this theorem, we only need to calculate the $\lambda_i$ modulo $p$. In our case, of all $\kappa_i$ and $f_i$ being linear modulo $p$, we get as special case the following corrollary.
\begin{cor}
\label{cor:finedixlinear}
If the system of equations $\forall i: \lambda_i \equiv 0 \bmod p$ has respectively no or one solution, there is respectively none or at most one point in $C(\Z)_P$.
\end{cor}

In general, even if $\overline{A}/\overline{I}$ is not finite, we see $\Hom(A/I,\Z_p)$ factors through $A/(I:p)$ where $(I:p) = \{x \in A \mid \exists k \in \Z_{\geq 0} p^kx \in A\}$. Then a higher precision calculation of the $\lambda_i$ can still lead to a succesfull application of Proposition~\ref{prop:finedix}.

Another specific case is that of $r = 1$. In that case, we can use finite precision approximations of $\lambda_1$ to deduce information about its Newton polygon, and use that to bound the number of zeroes in $\Z_p$. We can even adapt this method if $r$ is bigger than $1$; sometimes it might be possible to use the implicit function theorem for powerseries, Lemma~\ref{lem:implicitfunction}, to write one of the variables as a power series in the other variables, and then substitute it in the other equations.

\begin{exmp}
If $r = g-1 = 2$ and $\lambda_1 \equiv px-py,\lambda_2 \equiv x+y+pxy \bmod p^2$, neither Proposition~\ref{prop:finedix} nor Newton polygons are instantly applicable. However, substituting $y = -x +px^2 \bmod p^2$ into the first equation gives $2p(x-x^2) = 0 \bmod p^2$, or $x-x^2 = 0\bmod p$. Now both Proposition~\ref{prop:finedix} and Newton polygons give an upper bound of $2$ for $C(\Z)_P$.
\end{exmp}

@@

We first calculate $\Lcal(2D_0 - D_{\text{aff}})$. Both over $\Q_p$ and over $\F_p$ we have the equality
\[
\Lcal(2D_0 - D_{\text{aff}}) = a(x)\Lcal((2g+2-d)(\infty_- + \infty_+)) + (y-b(x))\Lcal((g+1)(\infty_- + \infty_+))
\]
and then it should follow over $\Z_p$ so over $\Z/p^k\Z$. \todo{Give more details, currently talking with Mascot about this.}

@@

\subsection{Parameters at $J$}
\label{subs:paramj}
Being able to calculate in the Jacobian, we can move on to the final ingredient for explicit computations: parameters at points of $J$. For any nice curve $C$ over $\Z$, smooth over $\Z_{(p)}$, with a $\Z$-point $b$, we have a birational map from $C^{(g)}$, the $g$-fold symmetric product of $C$, to $J$, given on points by sending $[(P_1,\dots,P_g)]$ to $[\sum P_i - gb]$. This map is \'etale at $[(P_1,\dots,P_g)]$ if $H^0(P_1 + \cdots + P_g)$ has dimension $1$ \todo{find a reference for this}, and if all $P_i$ are distinct, then the map $C^g \to C^{(g)}$ is \'etale at $P_1,\dots,P_g$ as well. Then finding parameters $t_1,\dots,t_g$ at $Q := [\sum_{i=1}^g P_i - gb]$ just comes down to finding parameters at each of the $P_i$. In our case of a hyperellptic curve, this is just an easy computation; for a point $(a,b)$ on the first affine chart, we can take $x-a$ if $b$ is non-zero and $y$ if $a$ is zero. We can also compute the inverse of the map $C(\Z/p^k\Z)_{P_i} \to p\Z/p^k\Z$; this is just a simple exercise in Hensel lifting.

This now means $J(\Z_p)_Q$ is parametrised as the product of $C(\Z_p)_{P_i}$. In particular, we have a bijection
\[
\prod_{i=1}^g C(\Z/p^k\Z)_{P_i} \to J(\Z/p^k\Z)_{Q}.
\] As we are able to explicitly compute this map using the Abel-Jacobi map, and we are able to test equality in $J(\Z/p^k\Z)$, we can compute the map $(t_1,\dots,t_g): J(\Z/p^k\Z)_Q \to (p\Z/p^{k}\Z)^g$ by first computing its inverse and storing all found values.

\label{}
The $\lambda_i$ from are in general not computationally available, as they are power series, but we can approximate them modulo powers of $p$. For example, we know that the $\lambda_i$ are linear modulo $p$. And in particular, if $f_1,\dots,f_{g-1}$ are as in Remark~\ref{rem:fislinear}, then $\lambda_i = \kappa_i$ for $i \leq g-1$, so they are also of degree at most $m$ modulo $p^m$ for $m < p-1$. Knowing this, we can compute $\lambda_i$ modulo powers of $p$ by extrapolation, and the following lemma, an immediate consequence of Lagrange interpolation, tells us how.
\begin{lem}
\label{lem:interpolate}
Let $R$ be a ring, and assume operations in $R$ are $O(1)$, $f \in R[x]$ with $x = (x_1,\dots,x_n)$ a polynomial of degree at most $m$, and $S$ a subset of $R$ with $S-S \subset \{0\} \cup R^*$ and $|S| > m$. Then $f$ is uniquely determinable in $O(nm^{n+1})$ by its values on $S^n$.
\end{lem}
\begin{proof}
Both uniqueness and construction follow from the case with $n=1$; uniqueness from the fact that the ideals $\{(x-s) \mid s \in S\}$ are pairwise coprime, and the construction is Lagrange interpolation. The complexity follows easily from induction on $n$ and the formula for Lagrange interpolation.
\end{proof}
